### 2021-05-20 concurrency/docs/theory/theory.tex
fix typo
write a more formal proof of the soundness of the TPCM-refinement borrowing concept

\newcommand{\hoare}[3]{\ensuremath{\{ #1 \}~ #2 ~\{ #3 \}}}
\newcommand{\viewshift}{\Rrightarrow}
\newcommand{\unit}{\epsilon}
\newcommand{\true}{\textsf{True}}
\newcommand{\false}{\textsf{False}}
\newcommand{\bool}{\textsf{bool}}
\DeclareMathOperator{\rel}{rel}
\DeclareMathOperator{\PCM}{PCM}
\DeclareMathOperator{\MonoidBase}{MonoidBase}
\DeclareMathOperator{\MonoidExt}{MonoidExt}
\DeclareMathOperator{\Frac}{Frac}
\newcommand{\monoid}[1]{\textsf{monoid}~#1}
\renewcommand{\eqdef}{\triangleq}
\newcommand{\borrow}[2]{\textsc{\&}^{#1}#2}
\newcommand{\pointsto}{\hookrightarrow}

@@

@@

We start by defining what it means to be a PCM with a transition relation $\mapsto$.

@@

We say that $\mapsto$ is \emph{monotonic} if
\[ \forall a, b, c.~ (a \mapsto b) \land (a \cdot c \ne \bot) \implies (b \cdot c \ne \bot) \land (a \cdot c \mapsto b \cdot c) \]
This allows us to define what it means to be a PCM,
\[
  \einfer{PCM}{
    \stackthree{
      \widehat{M} = (M, \cdot, \mapsto)
    }{
      \cdot : M \times M \to M^?, \text{~~comm., assoc.}
    }{
      \mapsto : M \times M \to \bool, \text{~~trans., refl., monotonic}
    }
  }{
    \PCM(\widehat{M})
  }
\]
Note that we can always construct a PCM by picking $\cdot$ to be commutative and associative,
and then by definition setting $\mapsto$ to be,
\[
  a \mapsto b = \forall c .~ a \cdot c \ne \bot \implies b \cdot c \ne \bot
\]
However, we can also set $\mapsto$ to be more restrictive, which will be useful later.

\textbf{PCM Example}

@@

We can define a PCM to correspond to this SSM by,
\[M \eqdef \{m \in S : \exists c .~ inv(m \cdot c) \}\]
\[a \cdot_M b \eqdef 
  \begin{cases}
      a \cdot b & \text{if } a \cdot b \in M\\
      \bot,              & \text{otherwise}
  \end{cases}\]
and finally we can define $\mapsto$ as the transitive closure of $\mapsto_\text{pre}$ defined as,
\[
a \mapsto_{\text{pre}} b \eqdef
  \exists a_0, b_0, c .~ (a = a_0 \cdot c) \land (b = b_0 \cdot c) \land a_0 \rightsquigarrow b_0
\]

\textbf{PCM refinements (motivation)}

@@

We'll develop the concept of a PCM refinement, denoted $\widehat{M} \refines \widehat{N}$.
In a modular design, $\widehat{M}$ will be the ``sub-component'', while $\widehat{N}$
is the client.

@@

\textbf{PCM refinements}

@@

\[
  \einfer{PCM-Refinement}{
    \stackfive{
      \PCM(\widehat{M}) \quad \PCM(\widehat{N})
    }{
      \rel : M \times N \to \bool
    }{
      \rel(\unit, \unit)
    }{
       (\forall b, b', q .~ b \mapsto b' \land \rel(b, q) \implies \exists q' .~ \rel(b', q') \land (q \mapsto q'))
    }{
      \forall b, a, a' .~ \rel(b,a) \land \rel(b,a') \implies a \mapsto a'
    }
  }{
    \widehat{M} \refines_{\rel} \widehat{N}
  }
\]

@@

Here are our rules for manipulating PCM ghost state, which are mostly standard except
the ones involving $\refines$.

@@

\[
  \einfer{Ghost-Alloc}{
    \PCM(\widehat{M}) \quad a\in M
  }{
    \true \viewshift \exists\alpha.~ \dboxed{a}^\alpha
  }
\]

\[
  \einfer{Ghost-Refinement-Alloc}{
    \stacktwo{
      \widehat{M} \refines_{\rel} \widehat{N}
    }{
      \rel(m, n) \quad b \ne \bot
    }
  }{
    \dboxed{n}^\gamma \viewshift \exists \alpha.~ \dboxed{m}^{\alpha \refines_{\rel} \gamma}
  }
\]

@@

Next, let's add some rules for making use of borrowed PCM state.

@@

Finally, rules to allow borrows to interact with PCM refinements:

@@

Note again in the second bullet that within the ``refinement worlds'' we allow both
\emph{borrowed and non-borrowed} state. This is the ``trick'' that lets the
PCM refinement be so effective for dealing with borrowed state.

@@

>>>>

\newcommand{\hoare}[3]{\ensuremath{\{ #1 \}~ #2 ~\{ #3 \}}}
\newcommand{\viewshift}{\Rrightarrow}
\newcommand{\unit}{\epsilon}
\newcommand{\true}{\textsf{True}}
\newcommand{\false}{\textsf{False}}
\newcommand{\bool}{\textsf{bool}}
\DeclareMathOperator{\rel}{rel}
\DeclareMathOperator{\TPCM}{TPCM}
\DeclareMathOperator{\MonoidBase}{MonoidBase}
\DeclareMathOperator{\MonoidExt}{MonoidExt}
\DeclareMathOperator{\Frac}{Frac}
\newcommand{\monoid}[1]{\textsf{monoid}~#1}
\renewcommand{\eqdef}{\triangleq}
\newcommand{\borrow}[2]{\textsc{\&}^{#1}#2}
\newcommand{\pointsto}{\hookrightarrow}

@@

\newcommand\Bigcircle{\raisebox{-0.5mm}{\scalebox{1.7}{$\bigcirc$}}}

@@

We start by defining what it means to be a PCM with a transition relation $\mapsto$.
We call this a \emph{Transitional PCM} (TPCM).

@@

We say that $\mapsto$ is \emph{monotonic} if
\[ \forall a, b, c.~ (a \mapsto b) \land (a \cdot c \ne \bot) \implies (b \cdot c \ne \bot) \land (a \cdot c \mapsto b \cdot c) \]
This allows us to define what it means to be a TPCM,
\[
  \einfer{TPCM}{
    \stackthree{
      \widehat{M} = (M, \cdot, \mapsto)
    }{
      \cdot : M \times M \to M^?, \text{~~comm., assoc.}
    }{
      \mapsto : M \times M \to \bool, \text{~~trans., refl., monotonic}
    }
  }{
    \TPCM(\widehat{M})
  }
\]
Note that we can always construct a TPCM by picking $\cdot$ to be commutative and associative,
and then by definition setting $\mapsto$ to be,
\[
  a \mapsto b = \forall c .~ a \cdot c \ne \bot \implies b \cdot c \ne \bot
\]
However, we can also set $\mapsto$ to be more restrictive, which will be useful later.

\textbf{TPCM Example}

@@

We can define a TPCM to correspond to this SSM by,
\[M \eqdef \{m \in S : \exists c .~ inv(m \cdot c) \}\]
\[a \cdot_M b \eqdef 
  \begin{cases}
      a \cdot b & \text{if } a \cdot b \in M\\
      \bot,              & \text{otherwise}
  \end{cases}\]
and finally we can define $\mapsto$ as the transitive closure of $\mapsto_\text{pre}$ defined as,
\[
a \mapsto_{\text{pre}} b \eqdef
  \exists a_0, b_0, c .~ (a = a_0 \cdot c) \land (b = b_0 \cdot c) \land a_0 \rightsquigarrow b_0
\]

\textbf{TPCM refinements (motivation)}

@@

We'll develop the concept of a TPCM refinement, denoted $\widehat{M} \refines \widehat{N}$.
In a modular design, $\widehat{M}$ will be the ``sub-component'', while $\widehat{N}$
is the client.

@@

\textbf{TPCM refinements}

@@

\[
  \einfer{TPCM-Refinement}{
    \stackfive{
      \TPCM(\widehat{M}) \quad \TPCM(\widehat{N})
    }{
      \rel : M \times N \to \bool
    }{
      \rel(\unit, \unit)
    }{
       (\forall b, b', q .~ b \mapsto b' \land \rel(b, q) \implies \exists q' .~ \rel(b', q') \land (q \mapsto q'))
    }{
      \forall b, a, a' .~ \rel(b,a) \land \rel(b,a') \implies a \mapsto a'
    }
  }{
    \widehat{M} \refines_{\rel} \widehat{N}
  }
\]

@@

Here are our rules for manipulating TPCM ghost state, which are mostly standard except
the ones involving $\refines$.

@@

\[
  \einfer{Ghost-Alloc}{
    \TPCM(\widehat{M}) \quad a\in M
  }{
    \true \viewshift \exists\alpha.~ \dboxed{a}^\alpha
  }
\]

\[
  \einfer{Ghost-Refinement-Alloc}{
    \stacktwo{
      \widehat{M} \refines_{\rel} \widehat{N}
    }{
      \rel(m, n)
    }
  }{
    \dboxed{n}^\gamma \viewshift \exists \alpha.~ \dboxed{m}^{\alpha \refines_{\rel} \gamma}
  }
\]

@@

Next, let's add some rules for making use of borrowed TPCM state.

@@

Finally, rules to allow borrows to interact with TPCM refinements:

@@

Note again in the second bullet that within the ``refinement worlds'' we allow both
\emph{borrowed and non-borrowed} state. This is the ``trick'' that lets the
TPCM refinement be so effective for dealing with borrowed state.

@@

\textbf{Justification (more formally)}

We can define a resource algebra (RA) as follows. Let $\Gamma$ be the set of locations
$\gamma$ and $K$ the set of lifetimes $\kappa$.
\begin{itemize}
  \item \emph{live} state at each location $\gamma$ given by $live : \Gamma \to m$
  \item \emph{reserved} state at each location $\gamma$ and lifetime $\kappa$,
      given by $reserved : K \to \Gamma \to m$.
  \item A set of \emph{borrows} where $borrows \subseteq K \times \Gamma \times M$.
  \item An active set of lifetimes, given by $active : K$.
\end{itemize}

The live state is additive (that is, $live(\gamma \to m) \cdot live(\gamma \to p) \eqdef live(\gamma \to p \cdot q)$.
Reserved state is similarly additive.

Finally, borrows are freely duplicable, and add \emph{freely} (i.e., take the union
of the set of borrow objects).

We'll describe the invariants on this resource algebra below, but first, I'll explain how
all the above concepts are embedded into this resource algebra:
\begin{align*}
 \dboxed{m}^\gamma &\eqdef live(\gamma \to m)\\
 \borrowed(\kappa, \dboxed{m}^\gamma) &\eqdef reserved(\kappa,\gamma \to m)\\
 \borrow{\kappa}{\dboxed{m}^\gamma} &\eqdef borrow(\kappa, \gamma \to m)\\
 \active(\kappa) &\eqdef active(\kappa)
\end{align*}

Now we define some functions on this state, defined recursively.
First, for any refinement location $\delta \refines_{\rel} \gamma$,
where $\rel : M \times N \to \bool$,
we let $Project_\delta : M \to N$ be a ``functionalization'' of $\rel$. In particular,
\[
  \forall m .~ domain(\rel, m) \implies \rel(m, Project_\delta(m)).
\]
where $domain(\rel, m) \eqdef \exists n .~ \rel(m, n)$.

Now we define $Total_\kappa(\gamma)$ to be the ``total'' state of a location $\gamma$:
that is, all the live and reserved state (for the active livetime $\kappa$)
\emph{plus} all the total state of locations which depend on $\gamma$. Thus we use
this recursive definition,

\[
  Total_\kappa(\gamma) \eqdef live(\gamma) \cdot
      \left(
        \Bigcircle_{\kappa' \ltGe \kappa} reserved(\kappa',\gamma)
      \right) \cdot
      \left(
        \Bigcircle_{\delta} Project_\delta\left(Total_\kappa(\delta \refines_{\rel} \gamma)\right)
      \right)
\]

Now, this definition is only sensible when 
$Total_\kappa(\delta \refines_{\rel} \gamma)$ is in the ``domain'' of $\rel$.

Thus we will have an invariant, only applying to the active lifetime $active$,
\begin{align}
  domain(rel, Total_{active}(\delta \refines_{\rel} \gamma)) \label{invTotal}
\end{align}

Next, we define a similar recursive function called $View$, which will be defined
only in terms of the $reserved$ state. However, it will be used to explain what the ``possible''
values of the $Total$ (minus the live state). A view will be a function $v : M \to \bool$.
These views will also be closed \emph{upwards}, i.e., $v(a) \implies v(a \cdot b)$.

Again, we define a ``projection'' function from a view on a refinement TPCM to a base TPCM.
\[
  ProjectView_\delta(v) \eqdef \lambda a .~
      \exists x, y .~ v(x) \land rel(x, y) \land y \le a.
\]
For a given $m$, we define $ViewOver(m)$ to be the view of all elements that are at least $m$:
\[
  ViewOver(m) \eqdef \lambda n .~ n \ge m
\]
And we can add two views in the natural way:
\[
  v \otimes w \eqdef \lambda m .~ \exists x, y .~ v(x) \land w(y) \land x \cdot y = m
\]
Finally, we're ready to define the $View_\kappa$ function for a given location, again
recursively, like the $Total_\kappa$ function:
\[
  View_\kappa(\gamma) \eqdef 
      \left(
        \bigotimes_{\kappa' \ltGe \kappa} ViewOver(reserved(\kappa,\gamma))
      \right) \otimes
      \left(
        \bigotimes_{\delta} ProjectView_\delta\left(View_\kappa(\delta \refines_{\rel} \gamma)\right)
      \right)
\]
Now the $View_\kappa$ denotes the ``possible'' states that the total unlive state can be in
for a given $\gamma$, without actually depending on any live state even in the refining locations. The $View_\kappa$ is in some sense used to justify the borrows at $\kappa$, so we will have an invariant,
\begin{align}
  & \forall \kappa, \gamma, m .~
  (\kappa, \gamma, m) \in borrows \implies \forall y .~ View_\kappa(\gamma)(m)(y) \implies m \le y
  \label{invView}
\end{align}
    %\left(
    %    \Bigcircle_{\kappa' \ltGe \kappa} reserved(\kappa,\gamma)
    %  \right) \cdot
    %  \left(
    %    \Bigcircle_{\delta} Project_\delta\left(Total_\kappa(\delta \refines_{\rel} \gamma)\right)
    %  \right)
   %\right)
%\end{align}
(We don't need a condition that $\kappa \ltGe active$, although we won't be able to make use
of the invariant otherwise.)

Now, we define $Inv$ to be both (1) and (2) together, and define an RA with a validity predicate $\mathcal{V}(x) \eqdef \exists y.~ Inv(x \cdot y)$. Then we can derive all the rules
claimed above, by simply showing that they are frame-preserving within this RA.

First, the borrowing rules: we can check that they preserve (\ref{invView}).
Also, all of the borrowing rules will preserve $Total_{active}$,
so they preserve (\ref{invTotal}) trivially.

Now, for the frame update rules (Borrow-Ghost-ViewShift and Borrow-Ghost-Refinement-Exchange).
First, these don't affect $View_{\kappa}$ at all, so it is easy to show that they preserve
(\ref{invView}).

We can also check that (\ref{invTotal}) is preserved. There are two key points here.
First, the rules about TPCM refinements imply that a frame-preserving update on a
$Total_\kappa(\delta \refines_{\rel} \gamma)$ implies a frame-preserving update
on the projection,
$Project_\delta(Total_\kappa(\delta \refines_{\rel} \gamma))$.

Second, we have to show why it is allowed to use borrowed state to justify the transitions.
Luckily, from (\ref{invTotal}) we can derive this law,
\[
  \kappa \ltGe active \implies View_\kappa(\gamma)(y)\left(
      \left(
        \Bigcircle_{\kappa' \ltGe active} reserved(\kappa,\gamma)
      \right) \cdot
      \left(
        \Bigcircle_{\delta} Project_\delta\left(Total_{active}(\delta \refines_{\rel} \gamma)\right)
      \right)
    \right).
    \label{invDerived}
\]
This large expression inside the $View_\kappa(\gamma)$ is just the part of
$Total_{active}(\gamma)$ except the part \emph{which is live at $\gamma$}.
Thus, combined with \ref{invView}, we can see that any borrow will be covered by the reserved 
state at $\gamma$ plus all state in the refined locations.

Just to reiterate the ``trick'': $View_\kappa(\gamma)$ depends only on reserved state:
however we are able to show that the value \[\left(
        \Bigcircle_{\kappa' \ltGe active} reserved(\kappa',\gamma)
      \right) \cdot
      \left(
        \Bigcircle_{\delta} Project_\delta\left(Total_{active}(\delta \refines_{\rel} \gamma)\right)
      \right)\] is in the active view, even though it includes the \emph{total} state at locations refining $\gamma$.
